Are populations like a circuit? The relationship between isolation by distance and isolation by resistance

### 2018-10-19 isolation_by_coalescence.tex
addressed comments mostly

In this paper, we contrast 
coalescence time and resistance distance,
and develop an inference method for movement rates on a discrete landscape
using both as estimated from genetic distances.
As we will see,
although the two methods are conceptually similar,
commute time inference may get the wrong answer if data derive from a coalescent process,
especially in the presence of asymmetry.
We also explore the question of identifiability.
Resistance-based methods are often used to predict distances based on a few layers of geographical data
(land cover, slope, etcetera).
Some methods find the combination of landscape layers to best explain genetic distance,
which allows very fine geographic resolution \citep{shaffer2017desert}.
Other methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information;
however, the geographical resolution is much coarser.
A recently-released method 
that is based on EEMS but uses haplotype length data and an underlying model of coalescence time
\citep{alasadi2018estimating}
is promising, but still produces similarly coarse maps.
One reason for this difference in resolution
is that EEMS uses more computationally intensive Bayesian methods.
However, we show that this is a more general problem,
both because of unmodeled ``process noise'' involved in discretization of geographical populations,
and the mathematical structure of the underlying problem.

@@

\paragraph{Constraints and uniqueness}
We have seen that to find the movement rates of a continuous-time Markov chain
given the coalescence or commute times
we must solve equations \eqref{eqn:C_matrix} or \eqref{eqn:R_from_H} and \eqref{eqn:G_from_H} for $G$,
and that these do not have unique solutions.
The situation is worse when one needs to infer coalescence rates ($\gamma$) as well.
However, usually in applications the locations lie across geographical space,
so that many of the entries of $G$ can be assumed to be zero.
This reduces the number of unknowns to solve for,
and so can render the solution unique.
We can get an idea of this by simply counting equations and unknowns.
For concreteness, suppose that the spatial locations
are arranged in a square grid of $n$ locations,
so that each is connected to four others.
(Boundary locations will have fewer, but we omit this detail.)
Since movement rates in each direction can be different,
there are $4n$ free parameters, each corresponding to an off-diagonal entry of $G$.
Coalescence rates provide another $n$ parameters.

@@

\paragraph{When are coalescence and commute times equal?}
If in a given biological situation commute and coalescence times are equal --
i.e., if given $G$ and $\gamma$, there was a $q$ that made $\comdist = C$ --
then inference with either model would be equivalent.
This was demonstrated in some situations by \citet{mcrae2006isolation}.
We have already shown that this is not the case in general,
but it turns out that it \emph{is} true
under some restrictive assumptions not uncommonly found in abstract population genetics models.
We show in Appendix \ref{sec:com_eq_coal} that
if hitting times are symmetric ($H = H^T$),
then this does occur,
and so the two methods are equivalent for symmetric island models
(where the populations are arranged on a ring 
and migration rates only depend on the distance between them).
Hitting times are not symmetric for a square grid with uniform migration
(the mean time to hit the center from a corner is less than the reverse),
but the square grid is quite close to the torus, which does have symmetric hitting times.

@@

\plr{transition paragraph}

@@

\paragraph{The effect of history}
The models we discuss here assume that population sizes and migration rates
have been constant on the time scale given by the within-species coalescent time.
This is rarely true in practice \citep{neigel1991estimation,barton1995genealogies}.
However, geographic differences in mean relatedness
are established on a shorter time scale --
the time scale over which a lineage, moving randomly across the landscape,
``forgets'' where it started
(i.e., the mixing time of the Markov chain).
This is the time it takes for the uncertainty in lineage movement
to reach the scale of the species range.
For instance,
if the width of a species' range is roughly 500km,
and if mean dispersal distance is 10km,
the standard deviation of ancestor location $t$ generations ago is $10\sqrt{t}$,
so the mixing time of a lineages across the landscape,
not accounting for barriers, is of order 2500 generations,
or about 50,000 years.
The landscape of many areas has changed substantially over this time, % \citep{pleistocene_mojave},
which suggests that models incorporating change over time 
may be required to model modern tortoise diversity.
However, the effects of recent history are strongest,
so landscapes estimated assuming constant populations
may give a reasonable picture of the landscape averaged over recent times.

@@

Thanks to David Levin for useful suggestions regarding hitting time calculations,
to Paul Marjoram for useful comments,
and to Brad Shaffer and Evan McCartney-Melstad for useful discussions
and providing the genetic distance data for desert tortoises.
We would also like to gratefully acknowledge the immense positive impact that Brad McRae \citep[1966--2017;][]{lawler2018tribute}
has had on landscape genetics and conservation biology through the introduction of landscape resistance.
We hope this paper helps to carry forward this work.
\plr{ERIK: also John, Hossein, Desi, ???}

>>>>

In this paper, we contrast 
coalescence time and resistance distance,
and develop an inference method for movement rates on a discrete landscape
using both as estimated from genetic distances.
As we will see,
although the two methods are conceptually similar,
commute time inference may get the wrong answer if data are derived from a coalescent process,
especially in the presence of asymmetry.
We also explore the question of identifiability.
Resistance-based methods are often used to predict distances based on a few layers of geographical data
(land cover, slope, etcetera).
Some methods find the combination of landscape layers to best explain genetic distance,
which allows very fine geographic resolution \citep{shaffer2017desert}.
Other methods such as EEMS \citep{petkova2016visualizing} infer conductances without such prior information;
however, the geographical resolution is much coarser.
A recently-released method 
that is based on EEMS but uses haplotype length data and an underlying model of coalescence time
\citep{alasadi2018estimating}
is promising, but still produces similarly coarse maps.
One reason for this difference in resolution
is that EEMS uses more computationally intensive Bayesian methods.
However, we show that this is a more general problem,
both because of unmodeled ``process noise'' involved in discretization of geographical populations,
and the mathematical structure of the underlying problem.

@@

\paragraph{Constraints and uniqueness}
We have seen that to find the movement rates of a continuous-time Markov chain
given the coalescence or commute times
we must solve either equation \eqref{eqn:C_matrix} 
or equations \eqref{eqn:G_from_H} and \eqref{eqn:R_from_H} for $G$,
and that these do not have unique solutions.
The situation is worse when one needs to infer coalescence rates ($\gamma$) as well.
However, usually in applications the locations lie across geographical space,
so that many of the entries of $G$ can be assumed to be zero.
This reduces the number of unknowns to solve for,
and so can render the solution unique.
We can get an idea of this by simply counting equations and unknowns.
For concreteness, suppose that the spatial locations
are arranged in a square grid of $n$ locations,
so that each is connected to four others.
(Boundary locations will have fewer, but we omit this detail.)
Since movement rates in each direction can be different,
there are $4n$ free parameters, each corresponding to an off-diagonal entry of $G$.
Coalescence rates provide another $n$ parameters.

@@

\paragraph{When are coalescence and commute times equal?}
If in a given biological situation commute and coalescence times are equal --
i.e., if given $G$ and $\gamma$, there was a $q$ that made $C = \comdist$ --
then inference with either model would be equivalent.
This was demonstrated in some situations by \citet{mcrae2006isolation}.
We have already shown that this is not the case in general,
but it turns out that it \emph{is} true
under some restrictive assumptions not uncommonly found in abstract population genetics models.
We show in Appendix \ref{sec:com_eq_coal} that
if hitting times are symmetric ($H = H^T$),
then this does occur,
and so the two methods are equivalent for symmetric island models
(where the populations are arranged on a ring 
and migration rates only depend on the distance between them).
Hitting times are not symmetric for a square grid with uniform migration
(the mean time to hit the center from a corner is less than the reverse),
but the square grid is quite close to the torus, which does have symmetric hitting times.

@@

@@

\paragraph{The effect of history}
The models we discuss here assume that population sizes and migration rates
have been constant on the time scale given by the within-species coalescent time.
This is rarely true in practice \citep{neigel1991estimation,barton1995genealogies}.
However, geographic differences in mean relatedness
are established on a shorter time scale --
the time scale over which a lineage, moving randomly across the landscape,
``forgets'' where it started
(i.e., the mixing time of the Markov chain).
This is the time it takes for the uncertainty in lineage movement
to reach the scale of the species range.
For instance,
if the width of a species' range is roughly 500km,
and if mean dispersal distance is 10km,
the standard deviation of ancestor location $t$ generations ago is $10\sqrt{t}$,
so the mixing time of a lineages across the landscape,
not accounting for barriers, is of order 2500 generations.
The habitat of many species would have changed substantially over this time, % \citep{pleistocene_mojave}
which suggests that models incorporating change over time 
may be required to model modern diversity.
However, the effects of recent history are strongest,
so landscapes estimated assuming constant populations
may give a reasonable picture of the landscape averaged over recent times.

@@

Thanks to David Levin for useful suggestions regarding hitting time calculations,
to Paul Marjoram for useful comments,
to Brad Shaffer and Evan McCartney-Melstad for discussions about landscape modeling,
and to John Novembre, Hussein Al-Asadi and Benjamin Peter for help with EEMS
and general input.
We would also like to gratefully acknowledge the immense positive impact that Brad McRae \citep[1966--2017;][]{lawler2018tribute}
has had on landscape genetics and conservation biology through the introduction of landscape resistance.
We hope this paper helps to carry forward this work.
